---
title: "Chapter 5"
author: "Keith McNulty"
date: "29/03/2022"
output: html_document
---


```{r}
library(rethinking)
library(dagitty)
```

## Divorce vs Marriage Age

```{r}
library(rethinking)
data(WaffleDivorce)
d <- WaffleDivorce

# standardize
d$D <- standardize(d$Divorce)
d$M <- standardize(d$Marriage)
d$A <- standardize(d$MedianAgeMarriage)

# check sd of median marriage age
sd(d$MedianAgeMarriage)

```


Compute posterior:

```{r}
# simple linear model
m5.1 <- quap(
  alist(
    D ~ dnorm(mu, sigma), # divorce normall distributed around my with sd sigma
    mu <- a + bA*A, # mu is linearly related to A
    a ~ dnorm(0, 0.2), # intercept is approximately zero 
    bA ~ dnorm(0, 0.5), #slope is mostly between -1 and 1
    sigma ~ dexp(1) # sd
  ),
  data = d
)
```

Simulate D from the priors:

```{r}
set.seed(10)

prior <- extract.prior(m5.1)
mu <- link(m5.1, post = prior, data = list(A = c(-2, 2)))
plot(NULL, xlim = c(-2, 2), ylim = c(-2, 2))
for (i in 1:50) {
  lines(c(-2, 2), mu[i, ], col = col.alpha("black", 0.4))
}

```

Plot posterior:

```{r}
A_seq <- seq(from = -3, to = 3.2, length.out = 30) # range of A for mean computation
mu <- link(m5.1, data = list(A = A_seq))
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

# plot
plot(D ~ A, data = d, col = rangi2)
lines(A_seq, mu.mean, lwd = 2)
shade(mu.PI, A_seq)
```

## Divorce Rate vs Marriage Rate

```{r}
# check sd of marriage rate
sd(d$Marriage)

# so if the slope was 1, we'd expect that 4% higher marriage rate would mean almost 4% higher divorce rate - seems reasonable
```

Compute posterior:

```{r}
m5.2 <- quap(
  alist(
    D ~ dnorm(mu, sigma), # normal
    mu <- a + bM*M, #linear
    a ~ dnorm(0, 0.1), # approx 0
    bM ~ dnorm(0, 0.5), # between -1 and 1 mostly
    sigma ~ dexp(1) 
  ),
  data = d
)
```

Plot posterior:

```{r}
M_seq <- seq(-3, 3.2, length.out = 30)
mu <- link(m5.2, data = list(M = M_seq))
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)
plot(D ~ M, data = d, col = rangi2)
lines(M_seq, mu.mean, lwd = 2)
shade(mu.PI, M_seq)
```

## Drawing a DAG

```{r}
dag5.1 <- dagitty("dag{A -> M; A -> D; M -> D}")
coordinates(dag5.1) <- list(x = c(A = 0, D = 1, M = 2), y = c(A = 0, M = 0, D = 1))
drawdag(dag5.1)
```


```{r}
DMA_dag1 <- dagitty("dag{D <- A -> M -> D}")
DMA_dag2 <- dagitty("dag{D <- A -> M}")

impliedConditionalIndependencies(DMA_dag1)
```

```{r}
impliedConditionalIndependencies(DMA_dag2)
```


## Multiple linear model

```{r}
m5.3 <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bM*M + bA*A,
    a ~ dnorm(0, 0.02),
    bM ~ dnorm(0, 0.5),
    bA ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = d
)

precis(m5.3)
```

```{r, eval = FALSE}
plot(coefs_list, par = c("bA", "bM")) 

```

## Predictor residual plots ~  Age at Marriage -> Marriage Rate

```{r}
m5.4 <- quap(
  alist(
    M ~ dnorm(mu, sigma),
    mu <- a + bAM * A,
    a ~ dnorm(0, 0.2),
    bAM ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = d
)

mu <- link(m5.4)
mu_mean <- apply(mu, 2, mean)
mu_resid <- d$M - mu_mean

# plot marriage rate against age at marriage
A_seq <- seq(-3, 3.2, length.out = 30)
mu <- link(m5.4, data = list(A = A_seq))
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)
plot(M ~ A, data = d, col = rangi2)
lines(A_seq, mu.mean, lwd = 2)
shade(mu.PI, A_seq)


```

```{r}
d_resid <- cbind(d, resid = mu_resid)

# residual model
resid.mod <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + bR*resid,
    a ~ dnorm(0, 0.02),
    bR ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = d_resid
)

mu <- link(resid.mod, data = list(resid = d_resid$resid))
mu_mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

plot(d_resid$resid, d_resid$D)
lines(d_resid$resid, mu_mean, lwd = 2)
shade(mu.PI, d_resid$resid)

```

## Posterior prediction plots - Predicted vs Observed

```{r}
# create predictions from observed data
mu <- link(m5.3)
mu_mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

# simulate instances using original data
D_sim <- sim(m5.3, n = 10000)
D_PI <- apply(D_sim, 2, PI)

# now plot
# first predicted vs observed
plot(mu_mean ~ d$D, col = rangi2, ylim = range(mu.PI),
     xlab = "Observed divorce rate", ylab = "Predicted divorce rate")

# add perfect correlation line
abline(a = 0, b = 1, lty = 2)

# add simulated PIs
for (i in 1:nrow(d)) {
  lines(rep(d$D[i], 2), mu.PI[, i], col = rangi2)
}

```

## Counterfactual plots

```{r}
m5.3_A <- quap(
  alist(
    ## A -> D <- M
    D ~ dnorm(mu, sigma),
    mu <- a + bM*M + bA*A,
    a ~ dnorm(0, 0.02),
    bM ~ dnorm(0, 0.5),
    bA ~ dnorm(0, 0.5),
    sigma ~ dexp(1),
    ## A -> M
    M ~ dnorm(mu_M, sigma_M),
    mu_M <- aM + bAM*A,
    aM ~ dnorm(0, 0.02),
    bAM ~ dnorm(0, 0.5), 
    sigma_M ~ dexp(1)
  ),
  data = d
)
```

Simulate counterfactual plots A -> D:

```{r}
A_seq <- seq(-2, 2, length.out = 30)

# simulate D amd M from A
sim_dat <- data.frame(A = A_seq)
s <- sim(m5.3_A, data = sim_dat, vars = c("M", "D"))

# plot
plot(sim_dat$A, colMeans(s$D), ylim = c(-2, 2), type = "l",
     xlab = "Manipulated A", ylab = "Counterfactual D")
shade(apply(s$D, 2, PI), sim_dat$A)
mtext("Total counterfactual effect of A on D")

```

And A -> M:

```{r}
# plot
plot(sim_dat$A, colMeans(s$M), ylim = c(-2, 2), type = "l",
     xlab = "Manipulated A", ylab = "Counterfactual M")
shade(apply(s$M, 2, PI), sim_dat$A)
mtext("Total counterfactual effect of A on M")

```

What about M -> D?

```{r}
sim2_dat <- data.frame(M = seq(-2, 2, length.out = 30), A = 0)
s2 <- sim(m5.3_A, data = sim2_dat, vars = "D")

plot(sim2_dat$M, colMeans(s2), ylim = c(-2, 2), type = "l",
     xlab = "Manipulated M", ylab = "Counerfactual D")

shade(apply(s2, 2, PI), sim2_dat$M)
mtext("Total counterfactual effect of M on D")

```


## Masked relationship

Load milk data:

```{r}
library(rethinking)
data(milk)
d <- milk
str(d)
```

Standardize:

```{r}
d$K <- standardize(d$kcal.per.g)
d$N <- standardize(d$neocortex.perc)
d$M <- standardize(d$mass)
```

Run a bivariate linear regression with vague priors:

```{r}
dcc <- d[complete.cases(d$K, d$N, d$M), ]

m5.5_draft <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bN*N,
    a ~ dnorm(0, 1),
    bN ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = dcc
)
```

Plot priors:

```{r}
prior <- extract.prior(m5.5_draft)
xseq <- c(-2, 2)
mu <- link(m5.5_draft, post = prior, data = list(N = xseq))
plot(NULL, xlim = xseq, ylim = xseq)
for (i in 1:50) {
  lines(xseq, mu[i, ], col = col.alpha("black", 0.3))
}

```

Recalibrate priors:

```{r}
m5.5 <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bN*N,
    a ~ dnorm(0, 0.2),
    bN ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = dcc
)

```

Check again:

```{r}
prior <- extract.prior(m5.5)
xseq <- c(-2, 2)
mu <- link(m5.5, post = prior, data = list(N = xseq))
plot(NULL, xlim = xseq, ylim = xseq)
for (i in 1:50) {
  lines(xseq, mu[i, ], col = col.alpha("black", 0.3))
}

```

Check posterior:

```{r}
precis(m5.5)
```

Plot posterior:

```{r}
xseq <- seq(min(dcc$N) - 0.15, max(dcc$N) + 0.15, length.out = 30)
mu <- link(m5.5, data = list(N = xseq))
mu_mean <- apply(mu, 2, mean)
mu_PI <- apply(mu, 2, PI)
plot(K ~ N, data = dcc)
lines(xseq, mu_mean, lwd = 2)
shade(mu_PI, xseq)
```
Now a bivariate model for body mass:

```{r}
dcc$logmass <- standardize(log(dcc$mass))
m5.6 <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bM*logmass,
    a ~ dnorm(0, 0.2),
    bM ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = dcc
)

precis(m5.6)
```

Plot posterior:

```{r}
xseq <- seq(min(dcc$logmass) - 0.15, max(dcc$logmass) + 0.15, length.out = 30)
mu <- link(m5.6, data = list(logmass = xseq))
mu_mean <- apply(mu, 2, mean)
mu_PI <- apply(mu, 2, PI)
plot(K ~ logmass, data = dcc)
lines(xseq, mu_mean, lwd = 2)
shade(mu_PI, xseq)
```

Now both:

```{r}
m5.7 <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a + bN * N + bM * logmass,
    a ~ dnorm(0, 0.2),
    bN ~ dnorm(0, 0.5),
    bM ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = dcc
)

precis(m5.7)
```

Counterfactual plot, manipulating `logmass`, holding N = 0:

```{r}
xseq <- seq(min(dcc$logmass) - 0.15, max(dcc$logmass) + 0.15, length.out = 30)
mu <- link(m5.7, data = list(logmass = xseq, N = 0))
mu_mean <- apply(mu, 2, mean)
mu_PI <- apply(mu, 2, PI)
plot(NULL, xlim = range(dcc$logmass), ylim = range(dcc$K))
lines(xseq, mu_mean, lwd = 2)
shade(mu_PI, xseq)

```

Counterfactual plot, manipulating N, holding `logmass` = 0:

```{r}
xseq <- seq(min(dcc$N) - 0.15, max(dcc$N) + 0.15, length.out = 30)
mu <- link(m5.7, data = list(logmass = rep(0, 30), N = xseq))
mu_mean <- apply(mu, 2, mean)
mu_PI <- apply(mu, 2, PI)
plot(NULL, xlim = range(dcc$N), ylim = range(dcc$K))
lines(xseq, mu_mean, lwd = 2)
shade(mu_PI, xseq)

```

Markov equivalent DAGs:

```{r}
dags5.7 <- dagitty(
  "dag{M -> K <- N; M -> N}"
)

coordinates(dags5.7) <- list(x = c(M = 0, K = 1, N =2), y = c(M = 0, N = 0, K = 1))
drawdag(dags5.7)
```

```{r}
MElist <- equivalentDAGs(dags5.7)
drawdag(MElist)
```

## Categorical variables

Model kcals according to slade:

```{r}
d$clade_id <- as.integer(d$clade)
d$K <- standardize(d$kcal.per.g)

m5.9 <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- a[clade_id],
    a[clade_id] ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data = d
)

precis(m5.9, depth = 2)
```

